Mechanization of Separation in Generic Extensions

### 2019-03-19 ITP/easy-axioms.tex
Primeros pasos hacia ITP

The Union Axiom asserts that if $x$ is a set, then there exists
another set (the union of $x$) containing all the elements in each
element of $x$. The relativized version of Union asks to give a name
$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi_a)=\union a$.
Let $\tau$ be the name for $a$, i.e.\ $a=\val(G,\tau)$; 
\citet{kunen2011set} gives $\pi_a$ in terms of $\tau$:
\begin{align*}
  \pi_a = \{\langle\theta,p \rangle :  %&\in \dom(\union(\dom(\tau))) \times \PP : \\
\exists \langle\sigma,q\rangle  \in \tau .
 \exists r . \langle \theta,r\rangle \in \sigma \wedge
    p\leqslant r \wedge p \leqslant q \}
\end{align*}
Our formal definition is slightly different in order to ease the proof
of $\pi_a \in M$; as it is defined using Separation, so one needs to
define the domain of separation and also internalize the predicate as
a formula
\isa{union{\isacharunderscore}name{\isacharunderscore}fm}. Instead of
working directly with the internalized formula, we define a predicate
\isa{Union{\isacharunderscore}name{\isacharunderscore}body} and prove the equivalence between
\begin{center}
\isa{sats(M,union{\isacharunderscore}name{\isacharunderscore}fm,[P,leq,\isasymtau,x])}
\end{center}
and
\isa{Union{\isacharunderscore}Name{\isacharunderscore}body(P,leq,\isasymtau,x)}. The
definition of $\pi_a$ in our formalization is:
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\ Union{\isacharunderscore}name\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}Union{\isacharunderscore}name{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ \isanewline
\ \ \ \ {\isacharbraceleft}u\ {\isasymin}\ domain{\isacharparenleft}{\isasymUnion}{\isacharparenleft}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}{\isacharparenright}{\isacharparenright}\ {\isasymtimes}\ P\ {\isachardot}\isanewline
\ \ \ \  \ \ \ \ Union{\isacharunderscore}name{\isacharunderscore}body{\isacharparenleft}P{\isacharcomma}leq{\isacharcomma}{\isasymtau}{\isacharcomma}u{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}

@@

The proof of Infinity for $M[G]$ takes advantage of some absoluteness
results proved in the locale \isa{M{\isacharunderscore}trivial}; this
proof is easy because we work in the context of the locale
\isa{M{\isacharunderscore}extra{\isacharunderscore}assms} which states
the assumption $\chk(x) \in M$ whenever $x\in M$. Since we have
already proved that $M[G]$ is transitive, $\emptyset\in M[G]$ assuming
$G$ being generic, and also that it satisfies Pairing and Union, we
can instantiate \isa{M{\isacharunderscore}trivial}:
\begin{isabelle}
\isacommand{sublocale}\isamarkupfalse%
\ G{\isacharunderscore}generic\ {\isasymsubseteq}\ M{\isacharunderscore}trivial{\isachardoublequoteopen}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}
We assume that $M$ satisfies Infinity, i.e., that Infinity relativized
to $M$ holds; therefore we obtain $I \in M$ such that $\emptyset\in I$
and, $x \in I$ implies $\mathit{succ}(x)\in I$ by absoluteness of
empty and successor for $M$. Using the assumption that $M$ is closed
under $\chk$, we deduce $\val(G,\chk(I)) = I \in M[G]$.  Now we can
use absoluteness of emptyness and successor, this time for $M[G]$, to
conclude that $M[G]$ satisfies Infinity.
% \begin{isabelle}
%   \isacommand{lemma}\isamarkupfalse%
% \ infinty{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
% \end{isabelle}

>>>>

The Union Axiom asserts that if $x$ is a set, then there exists
another set (the union of $x$) containing all the elements in each
element of $x$. The relativized version of Union asks to give a name
$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi_a)=\union a$.
Let $\tau$ be the name for $a$, i.e.\ $a=\val(G,\tau)$; 
Kunen \cite{kunen2011set} gives $\pi_a$ in terms of $\tau$:
\begin{align*}
  \pi_a = \{\langle\theta,p \rangle :  %&\in \dom(\union(\dom(\tau))) \times \PP : \\
\exists \langle\sigma,q\rangle  \in \tau .
 \exists r . \langle \theta,r\rangle \in \sigma \wedge
    p\leqslant r \wedge p \leqslant q \}
\end{align*}
Our formal definition is slightly different in order to ease the proof
of $\pi_a \in M$; as it is defined using Separation, so one needs to
define the domain of separation and also internalize the predicate as
a formula
\isa{union{\isacharunderscore}name{\isacharunderscore}fm}. Instead of
working directly with the internalized formula, we define a predicate
\isa{Union{\isacharunderscore}name{\isacharunderscore}body} and prove the equivalence between
\begin{center}
\isa{sats(M,union{\isacharunderscore}name{\isacharunderscore}fm,[P,leq,\isasymtau,x])}
\end{center}
and
\isa{Union{\isacharunderscore}Name{\isacharunderscore}body(P,leq,\isasymtau,x)}. The
definition of $\pi_a$ in our formalization is:
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\ Union{\isacharunderscore}name\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}Union{\isacharunderscore}name{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ \isanewline
\ \ \ \ {\isacharbraceleft}u\ {\isasymin}\ domain{\isacharparenleft}{\isasymUnion}{\isacharparenleft}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}{\isacharparenright}{\isacharparenright}\ {\isasymtimes}\ P\ {\isachardot}\isanewline
\ \ \ \  \ \ \ \ Union{\isacharunderscore}name{\isacharunderscore}body{\isacharparenleft}P{\isacharcomma}leq{\isacharcomma}{\isasymtau}{\isacharcomma}u{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}

@@

The proof of Infinity for $M[G]$ takes advantage of some absoluteness
results proved in the locale \isa{M{\isacharunderscore}trivial}; this
proof is easy because we work in the context of the locale
\isa{M{\isacharunderscore}extra{\isacharunderscore}assms} which states
the assumption $\chk(x) \in M$ whenever $x\in M$. Since we have
already proved that $M[G]$ is transitive, $\emptyset\in M[G]$ assuming
$G$ being generic, and also that it satisfies Pairing and Union, we
can instantiate \isa{M{\isacharunderscore}trivial}:
\begin{isabelle}
\isacommand{sublocale}\isamarkupfalse%
\ G{\isacharunderscore}generic\ {\isasymsubseteq}\ M{\isacharunderscore}trivial{\isachardoublequoteopen}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}
We assume that $M$ satisfies Infinity, i.e., that Infinity relativized
to $M$ holds; therefore we obtain $I \in M$ such that $\emptyset\in I$
and, $x \in I$ implies $\mathit{succ}(x)\in I$ by absoluteness of
empty and successor for $M$. Using the assumption that $M$ is closed
under $\chk$, we deduce $\val(G,\chk(I)) = I \in M[G]$.  Now we can
use absoluteness of emptiness and successor, this time for $M[G]$, to
conclude that $M[G]$ satisfies Infinity.
% \begin{isabelle}
%   \isacommand{lemma}\isamarkupfalse%
% \ infinty{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
% \end{isabelle}

### 2019-03-19 ITP/intro.tex
Primeros pasos hacia ITP

In this work we address a substantial part of formalizing the proof
that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
obtained by forcing also satisfies $\ZF$. As remarked by
\citet[][p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$
  is a model for set theory, the hardest axiom to verify is
  [Separation].}  
The most important achievement of this paper is the
mechanization in the proof assistant \emph{Isabelle} of a proof of the
Axiom of Separation in 
generic extensions by using the ``fundamental'' theorems of forcing.
En route
to this, we also formalized the satisfaction by $M[G]$ of
Extensionality, Foundation, and Union. As a consequence of Separation
we were able to formalize the proof of the Powerset Axiom; finally,
the Axiom of Infinity was proved under extra assumptions. % We have already
% proved Pairing in the first report of our project
% \cite{2018arXiv180705174G}.
The theoretical support for this work has been 
the fine textbook by Kunen \cite{kunen2011set} and our development
benefited from the remarkable work done by Lawrence 
Paulson \cite{paulson_2003} on the formalization of G\"odel's
constructible universe in Isabelle.

@@

The \emph{Formal Abstracts} project~\cite{hales-fabstracts} proposes
the formalization of complex pieces of mathematics by writing the
statements of 
results and the material upon which they are based (definitions,
propositions, lemmas), but ommiting the proofs. In this work we
partially adhere to this vision to delineate our formalization
strategy:
Since the proofs that the  axioms hold in generic extensions
are independent of the \emph{proofs} of the fundamental theorems of
forcing, we assumed the latter for the time being. Let us remark
that those theorems depend on the definition of a function $\forceisa$
from formulas to formulas which is, by itself, quite demanding; the
formalization of it and of the fundamental theorems of forcing % roughly
comprises barely less than a half of our full project.

It might be a little surprising the lack of formalizations of forcing
and generic extensions. As far as we know, the development of
\citet{JFR6232} in homotopy type theory for constructing generic
extensions in a sheaf-theoretic setting is the unique mechanization of
forcing. This contrast with the fruitful use of forcing techniques to
extend the Curry-Howard isomorphism to classical axioms
\cite{Miquel:2011:FPT:2058525.2059614,lmcs:1070}. Moreover, the
combination of forcing with intuitionistic type theory
\cite{Coquand:2009:FTT:1807662.1807665,coquand2010note} gives rise
both to positive results (an algorithm to obtain witnesses of the
continuity of definable functionals \cite{coquand2012computational})
and also negative (the independence of Markov's principle
\cite{lmcs:3859}). In the same strand of forcing from the point of
view of proof theory \cite{avigad_2004} are the conservative
extensions of CoC with forcing conditions
\cite{jaber:hal-01319066,Jaber:2012:ETT:2358958.2359524}.

@@

In pursuing the proof of Separation on generic extensions we
extended Paulson's library with:
\begin{inlinelist}
\item renaming of variables for \emph{internalized} formulas, which
  with little effort can be extended 
  to substitutions;
\item an improvement on definitions by recursion on well-founded
  relations; 
\item enhancements in the hierarchy of locales; and
\item a variant of the  principle of dependent choices and a version
  of Rasiowa-Sikorski, which 
  ensures the existence of generic filters for countable and transitive
  models of $\ZF$;
\end{inlinelist} 
the last item was already communicated in the
  first report \cite{2018arXiv180705174G}.

We briefly describe the contents of each
section. Section~\ref{sec:isabelle} contains the bare minimun
requirements to understand the (meta)logics used in Isabelle. Next, an
overview of the model theory of set theory is presented in
Section~\ref{sec:axioms-models-set-theory}. There is an ``internal''
representation of first-order formulas as sets, implemented by
Paulson; Section~\ref{sec:renaming} discusses syntactical
transformations of the former, mainly permutation of variables. 
In Section~\ref{sec:generic-extensions} the generic extensions are
succintly reviewed and how the treatment of well founded recursion in
Isabelle was enhanced. We take care of the ``easy axioms'' in
Section~\ref{sec:easy-axioms}; these are the ones that
do not depend on the forcing theorems. We describe the latter in
Section~\ref{sec:forcing}. We adapted the  work by Paulson to our
needs, and this is described in
Section~\ref{sec:hack-constructible}. We present the proof
of the Separation Axiom Scheme in Section~\ref{sec:proof-separation},
which follows closely its implementation, and some comments on the
proof of the Powerset Axiom. A plan for future work and
some immediate conclusions are offered in
Section~\ref{sec:conclusions-future-work}.

>>>>

In this work we address a substantial part of formalizing the proof
that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
obtained by forcing also satisfies $\ZF$. As remarked by Kunen
\cite[p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$ is
  a model for set theory, the hardest axiom to verify is
  [Separation].}  The most important achievement of this paper is the
mechanization in the proof assistant \emph{Isabelle} of a proof of the
Axiom of Separation in generic extensions by using the ``fundamental''
theorems of forcing.  En route to this, we also formalized the
satisfaction by $M[G]$ of Extensionality, Foundation, and Union. As a
consequence of Separation we were able to formalize the proof of the
Powerset Axiom; finally, the Axiom of Infinity was proved under extra
assumptions. % We have already
% proved Pairing in the first report of our project
% \cite{2018arXiv180705174G}.
The theoretical support for this work has been 
the fine textbook by Kunen \cite{kunen2011set} and our development
benefited from the remarkable work done by Lawrence 
Paulson \cite{paulson_2003} on the formalization of G\"odel's
constructible universe in Isabelle.

@@

The \emph{Formal Abstracts} project~\cite{hales-fabstracts} proposes
the formalization of complex pieces of mathematics by writing the
statements of 
results and the material upon which they are based (definitions,
propositions, lemmas), but omitting the proofs. In this work we
partially adhere to this vision to delineate our formalization
strategy:
Since the proofs that the  axioms hold in generic extensions
are independent of the \emph{proofs} of the fundamental theorems of
forcing, we assumed the latter for the time being. Let us remark
that those theorems depend on the definition of a function $\forceisa$
from formulas to formulas which is, by itself, quite demanding; the
formalization of it and of the fundamental theorems of forcing % roughly
comprises barely less than a half of our full project.

It might be a little surprising the lack of formalizations of forcing
and generic extensions. As far as we know, the development of
Quirin and Tabareau \cite{JFR6232} in homotopy type theory for constructing generic
extensions in a sheaf-theoretic setting is the unique mechanization of
forcing. This contrast with the fruitful use of forcing techniques to
extend the Curry-Howard isomorphism to classical axioms
\cite{Miquel:2011:FPT:2058525.2059614,lmcs:1070}. Moreover, the
combination of forcing with intuitionistic type theory
\cite{Coquand:2009:FTT:1807662.1807665,coquand2010note} gives rise
both to positive results (an algorithm to obtain witnesses of the
continuity of definable functionals \cite{coquand2012computational})
and also negative (the independence of Markov's principle
\cite{lmcs:3859}). In the same strand of forcing from the point of
view of proof theory \cite{avigad_2004} are the conservative
extensions of CoC with forcing conditions
\cite{jaber:hal-01319066,Jaber:2012:ETT:2358958.2359524}.

@@

In pursuing the proof of Separation on generic extensions we
extended Paulson's library with:
\begin{enumerate}
\item renaming of variables for \emph{internalized} formulas, which
  with little effort can be extended 
  to substitutions;
\item an improvement on definitions by recursion on well-founded
  relations; 
\item enhancements in the hierarchy of locales; and
\item a variant of the  principle of dependent choices and a version
  of Rasiowa-Sikorski, which 
  ensures the existence of generic filters for countable and transitive
  models of $\ZF$;
\end{enumerate}
the last item was already communicated in the
  first report \cite{2018arXiv180705174G}.

We briefly describe the contents of each
section. Section~\ref{sec:isabelle} contains the bare minimum
requirements to understand the (meta)logics used in Isabelle. Next, an
overview of the model theory of set theory is presented in
Section~\ref{sec:axioms-models-set-theory}. There is an ``internal''
representation of first-order formulas as sets, implemented by
Paulson; Section~\ref{sec:renaming} discusses syntactical
transformations of the former, mainly permutation of variables. 
In Section~\ref{sec:generic-extensions} the generic extensions are
succinctly reviewed and how the treatment of well founded recursion in
Isabelle was enhanced. We take care of the ``easy axioms'' in
Section~\ref{sec:easy-axioms}; these are the ones that
do not depend on the forcing theorems. We describe the latter in
Section~\ref{sec:forcing}. We adapted the  work by Paulson to our
needs, and this is described in
Section~\ref{sec:hack-constructible}. We present the proof
of the Separation Axiom Scheme in Section~\ref{sec:proof-separation},
which follows closely its implementation, and some comments on the
proof of the Powerset Axiom. A plan for future work and
some immediate conclusions are offered in
Section~\ref{sec:conclusions-future-work}.

### 2019-03-19 ITP/renaming.tex
Primeros pasos hacia ITP

In the course of our work we need to reason about renaming of formulas
and its effect on their satisfiability. Internalized formulas are
implemented using de Bruijn indices for variables and the arity of a
formula $\phi$ gives the least natural number containing all the free
variables in $\phi$. Following \citet{fiore-abssyn}, one can
understand the arity of a formula as the context of the free
variables; notice that the arity of $\forallFm{\phi}$ is the
predecessor of the arity of $\phi$. Renamings are, consequently,
mappings between finite sets; since we can think of $\mathsf{succ}(n)$
as the coproduct $1+n = \{0\} \cup \{1,\dots,n\}$, then given a
renaming $f \colon n \to m$, the 
unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to rename
free variables in a quantified formula.

>>>>

In the course of our work we need to reason about renaming of formulas
and its effect on their satisfiability. Internalized formulas are
implemented using de Bruijn indices for variables and the arity of a
formula $\phi$ gives the least natural number containing all the free
variables in $\phi$. Following Fiore et al. \cite{fiore-abssyn}, one can
understand the arity of a formula as the context of the free
variables; notice that the arity of $\forallFm{\phi}$ is the
predecessor of the arity of $\phi$. Renamings are, consequently,
mappings between finite sets; since we can think of $\mathsf{succ}(n)$
as the coproduct $1+n = \{0\} \cup \{1,\dots,n\}$, then given a
renaming $f \colon n \to m$, the 
unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to rename
free variables in a quantified formula.

### 2019-03-19 ITP/zfc-axioms.tex
Primeros pasos hacia ITP

The axioms of Zermelo and Fraenkel ($\ZF$) form a
countably infinite list of first-order sentences in a language
consisting of an only binary relation symbol $\in$. These include the
axioms of Extensionality, Pairing, Union, Powerset, Foundation, Infinity, and two
axiom-schemes collectively referred as
\begin{inlinelist}[label=(\emph{\alph*})]
\item Axiom of Separation:
\emph{For every $A$, $a_1,\dots,a_n$,  and  a formula
  $\psi(x_0,x_1,\dots,x_n)$, there exists $\{a\in  A:
  \psi(a,a_1,\dots,a_n)\}$}, 
and \item Axiom of Replacement: \emph{For every $A$, $a_1,\dots,a_n$,  and
  a formula   $\psi(x,z,x_1\dots,x_n)$, if 
  $\forall x.\exists!z.\psi(x,z,x_1,\dots,x_n)$,  there exists 
  $\{b : \exists a\in A. \psi(a,b,a_1,\dots,a_n)\}$}.
\end{inlinelist}
An excellent introduction to the axioms and the motivation behind them
can be found in Shoenfield \cite{MR3727410}.

@@

The rest of the axioms are also
included. We single out Separation and Replacement: These are written
for formulas with at most one extra parameter (meaning $n\leq 1$ in the
above $\psi$). Thanks to Pairing, these 
versions are equivalent to the usual formulations. We are only able to
prove that the generic extension satisfies Separation for any particular
number of parameters, but not in general. This is a consequence that
induction on terms of type \tyo{} is not available.

>>>>

The axioms of Zermelo and Fraenkel ($\ZF$) form a
countably infinite list of first-order sentences in a language
consisting of an only binary relation symbol $\in$. These include the
axioms of Extensionality, Pairing, Union, Powerset, Foundation, Infinity, and two
axiom-schemes collectively referred as
(a) \emph{For every $A$, $a_1,\dots,a_n$,  and  a formula
  $\psi(x_0,x_1,\dots,x_n)$, there exists $\{a\in  A:
  \psi(a,a_1,\dots,a_n)\}$}, 
and (b) Axiom of Replacement: \emph{For every $A$, $a_1,\dots,a_n$,  and
  a formula   $\psi(x,z,x_1\dots,x_n)$, if 
  $\forall x.\exists!z.\psi(x,z,x_1,\dots,x_n)$,  there exists 
  $\{b : \exists a\in A. \psi(a,b,a_1,\dots,a_n)\}$}.
An excellent introduction to the axioms and the motivation behind them
can be found in Shoenfield \cite{MR3727410}.

@@

The rest of the axioms are also included. We single out Separation and
Replacement: These are written for formulas with at most one extra
parameter (meaning $n\leq 1$ in the formula $\psi$ of the
axiom-schemes above). Thanks to Pairing, these versions are equivalent
to the usual formulations. We are only able to prove that the generic
extension satisfies Separation for any particular number of
parameters, but not in general. This is a consequence that induction
on terms of type \tyo{} is not available.
